Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms

### 2013-05-31 CGP Journal/main.tex
Big block of results

\subsection{DAG}
The principle behind \emph{DAG} is to remove the traditional CGP requirement that
nodes can only take input from nodes that precede them in the genome.  Instead
this is replaced with the relaxed requirement than the genome cannot encode for
any cycles.  To accomplish this goal, modifications must be made to how CGP
performs mutation and evaluation.

@@

\subsection{Parameter Setting}
In order to fairly compare each method, we should ensure proper parameter configuration.
This is to avoid the potential criticism that the parameter settings chosen benefit
a specific method more than others.  While tuning can lead to the alternative issue
of methods only being effective after extensive problem specific tuning, we set out to use
rough grained values and provide each technique with equal tuning time to alleviate that bias.

@@

	\end{tabular}
	\caption{Multiply 5.1429910779824196e-18}
	\label{tab:Multiply}
\end{table*}

TODO Compare using a table the best parameters found for each configuration.
One table per problem.

\subsection{Scalability}
TODO Discuss using tuned parameters on different problem sizes and analyze
results.  Include O(N) table similar to previous paper.

>>>>

\subsection{DAG}
\label{sec:dag}
The principle behind \emph{DAG} is to remove the traditional CGP requirement that
nodes can only take input from nodes that precede them in the genome.  Instead
this is replaced with the relaxed requirement than the genome cannot encode for
any cycles.  To accomplish this goal, modifications must be made to how CGP
performs mutation and evaluation.

@@

\subsection{Parameter Setting}
\label{sec:parameter}
In order to fairly compare each method, we should ensure proper parameter configuration.
This is to avoid the potential criticism that the parameter settings chosen benefit
a specific method more than others.  While tuning can lead to the alternative issue
of methods only being effective after extensive problem specific tuning, we set out to use
rough grained values and provide each technique with equal tuning time to alleviate that bias.

@@

	\end{tabular}
	\caption{Multiply 5.1429910779824196e-18}
	\label{tab:multiply}
\end{table*}

The results from running each tuned parameter configuration for each algorithm
combination on each problem are summarized by Table~\ref{tab:parity},
Table~\ref{tab:encode}, Table~\ref{tab:decode}, and Table~\ref{tab:multiply}.
Rows in each table are ordered by that algorithm combination's median evaluations to
success (MES), with the best algorithm combination appearing at the top of each table.
The highlighted row marks \emph{Skip Normal}, the control algorithm combination
representing traditional CGP's performance.  All configurations for each problem
were then testing using Kruskal-Wallis to determine if any configuration is
statistically different from any other.  All problems received highly significant
values.  As such, each algorithm combination was then compared with the control
to determine statistical significance using pairwise Mann-Whitney~U tests. The
resulting p-value for each test is reported in the table.

As discussed in Section~\ref{sec:parameter}, the genome size and mutation rate
for each algorithm combination was extensively optimized.  In line with previous
research~\cite{miller:2006:redundancy}, this led to massive genome sizes for
almost all algorithm combinations on all problems.  With the exception of 
some combinations using \emph{Single}, all algorithm combinations used over
1,000 nodes.  Half of those configurations used 10,000 nodes, the maximum allowed
by our tuning, which may imply these combinations work best with even larger
genomes.

Of the 12 configurations including \emph{Single}, 11 used a genome size smaller
than all configurations not using \emph{Single}.  The exception was \emph{Single Normal}
on Multiply.  No other algorithm had as much impact on genome size tuning as
the inclusion or exclusion of \emph{Single}.  In previous work~\cite{goldman:2013:ordering}
it appeared \emph{Reorder} and \emph{DAG} were allowing CGP to use smaller genomes.
In light of our results, this conclusion should be amended.
It appears that \emph{Single} is most effective when the genome size is small and
that \emph{Reorder} and \emph{DAG} are best able to cope with smaller genome sizes.
Independent of duplication method, the ordering methods do not seem to create any
pressure on which genome size is chosen.

\begin{comment}
Genome size explosion seems most prevalent on easier problems, with Parity and
Encode having much higher genome sizes than Decode and Multiply.  Initially
this seems counter intuitive, as we would expect harder problems to require
larger solutions which in turn should require larger genomes.  Yet the genome
sizes in use are so far beyond what a solution to these problems require that
solution size is likely not the cause of this behavior.  Instead, consider
the usefulness of unexplored nodes.

On simple enough problems, exploring lots of randomly generated nodes may be
sufficient to generate high quality solutions.  At relatively low numbers of
evaluations, large genotypes likely mean each mutation to a connection gene
will activate nodes that have never been active, resulting in exploration.
At higher numbers of evaluations, mutations to connection genes that are not currently
active have a high chance of connecting to a node that was tested previously
and found ineffective.  As a result, having a smaller genome size on harder problems
may reduce the chance of exploratory mutations when optimizing near the global optimum.

Under this theory, we should expect Parity to have the largest genome sizes
\end{comment}

Unlike genome size, all configurations choose approximately the same mutation rate,
with different rates for each problem.  With the exception of Parity, there appears
to be a general consensus that each gene should be mutated with probability 0.002,
independent of genome size.  Parity used a higher mutation rate, but this is likely
because 3-bit Parity is by far the easiest problem tested and therefore likely benefits
from higher exploration.

On many of the problems, many of the configurations have little to no statistical
difference from the control in their evaluations required to reach the global optimum.
On Parity and Encode it is probably safe to conclude that \emph{Single} is worse than
the control, although as these are the easiest problems tested that difference may have
little real world meaning.  \emph{DAG} is almost certainly worse than the control on
Decode.  Its results are so much worse that we can also likely conclude that \emph{DAG}
is worse than all non \emph{DAG} configuration on that problem, even though no statistical
tests are presented here.  The control on multiply did by far the worst of all configurations,
with only \emph{Accumulate Normal} likely failing statistical significance.

By ordering the algorithm combinations by MES and examining that ordering across each
problem, we can look for general trends.  From the tables, it appears on easier problems
which method of duplication is in use is the strongest predictor of rank, while
the ordering methods are more important on harder problems.  This comes from the
observation that in for Parity and Encode (Table~\ref{tab:parity} and Table~\ref{tab:encode})
the first column is effectively sorted, while for Decode and Multiply
(Table~\ref{tab:decode} and Table~\ref{tab:multiply}) the second column is sorted.
Each table includes coincidentally one exception to these rules.  On Parity and
Encode pairings with \emph{DAG} are ranked lower than other ordering methods, so much
so as to break the ordering once on both problems.  On Decode \emph{Reorder Duplicate}
ranks worse than expected.  Finally on Multiply \emph{Single DAG} does far better
than expected.

On almost every problem, \emph{DAG} required
more evaluations to success than \emph{Reorder}, and it only outperformed \emph{Normal}
on Multiply.  Combined with its's algorithmic overhead described in
Section~\ref{sec:dag}, \emph{DAG} is therefore not likely to be useful for anything
but comparison purposes.

On all four problems, \emph{Reorder} either finds the solution in comparable
or significantly less evaluations than \emph{Normal} ordering, independent of
the duplicate method in use.  Combined with the complexity analysis in
Section~\ref{sec:reorder} which shows it is no more asymptotically complex,
we would suggest \emph{Reorder} as an alternative to \emph{Normal} on hard
problems.

The difference between using \emph{Skip} and \emph{Accumulate} remains
negligible~\cite{goldman:2013:cgpwaste}, even in combination with different ordering
methods.  As \emph{Accumulate} is far more complex in terms of algorithm function
than \emph{Skip}, we would suggest \emph{Skip} be used over \emph{Accumulate} for
any real world application.

TODO Discuss actual runtime analysis, and how \emph{Single Reorder} is likely
the fastest method in that regard.
\subsection{Scalability}
TODO Discuss using tuned parameters on different problem sizes and analyze
results.  Include O(N) table similar to previous paper.
